---
title: "Lab 02: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustainability 523c'
author:
  - name: Sierra Champion
    email: schamp03@colostate.edu
format: html
---

# Libraries

```{r}
install.packages('gghighlight')
install.packages('ggrepel')
install.packages('st_transform')

# spatial data science
library(tidyverse)
library(sf)
library(units)


# Data
library(AOI)
library(readr)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
library(remotes)
```

# Question 1

### **1.1 Define a Projection** 

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

boundaries_data<- remotes::install_github("mikejohnson51/AOI")

US_boundaries<- data.frame(boundaries_data)
```

## 1.2 Get State Boundaries

```{r}
library(AOI)
#Get USA State Boundaries
usa_states <- aoi_get(state = "conus") |>
  st_transform(crs = eqdc)

```

## 1.3 Get Country Boundaries

```{r}
#Get Country Boundaries
north_america <- aoi_get(country = c("MX", "CA", "USA")) %>%
  st_transform(crs = eqdc)

```

```{r}
remotes::install_github("ropensci/USAboundaries")
remotes::install_github("ropensci/USAboundariesData")
remotes::install_github("ropensci/rnaturalearthdata")
```

## 1.4 Get City Locations

```{r}
uscities <- read_csv("data/uscities.csv")

eqdc <- st_crs(5070)
states <- USAboundaries::us_states(resolution = "low") |>
  filter(!state_abbr %in% c("HI","AK","PR")) |>
  st_transform(eqdc)

countries <- rnaturalearth::countries110 |>
  st_transform(eqdc)


```

# Question 2

```{r}
# Use EPSG:5070 (Conus Albers)
crs_proj <- 5070

# Reproject everything
uscities_sf <- st_transform(uscities_sf, crs = crs_proj)
usa_border <- st_transform(usa_border, crs = crs_proj)

```

## 2.1 Distance to USA Border (km)

```{r}
library(tidyverse)
library(sf)
library(units)
library(flextable)

usa_border <- states |>
  st_union() |>                    
  st_cast("MULTILINESTRING")       

# calculate distance from each city to the USA border
uscities_sf <- uscities_sf |>
  mutate(dist_to_usa_border_km = set_units(st_distance(geometry, usa_border), "km"))

# 5 cities farthest from the USA border
uscities_sf |>
  st_drop_geometry() |>
  select(city, state_id, dist_to_usa_border_km) |>
  arrange(desc(dist_to_usa_border_km)) |>
  slice_head(n = 5) |>
  flextable()

```

## 2.2 Distance to States (km)

```{r}
state_borders <- states |>
  st_geometry() |>
  st_combine() |>
  st_cast("MULTILINESTRING")

uscities_sf <- uscities_sf |>
  mutate(dist_to_state_border_km = set_units(st_distance(geometry, state_borders), "km"))

uscities_sf |>
  st_drop_geometry() |>
  select(city, state_id, dist_to_state_border_km) |>
  arrange(desc(dist_to_state_border_km)) |>
  slice_head(n = 5) |>
  flextable()

```

## 2.3 Distance to Mexico (km)

```{r}

library(rnaturalearth)

countries <- ne_countries(scale = "small", returnclass = "sf")

mexico_border <- countries |>
  filter(sovereignt == "Mexico") |>
  st_union() |>
  st_cast("MULTILINESTRING") |>
  st_transform(5070)

uscities_sf <- uscities_sf |>
  mutate(dist_to_mexico_km = set_units(st_distance(geometry, mexico_border), "km"))

uscities_sf |>
  st_drop_geometry() |>
  select(city, state_id, dist_to_mexico_km) |>
  arrange(desc(dist_to_mexico_km)) |>
  slice_head(n = 5) |>
  flextable()

```

## Distance to Canada (km)

```{r}
uscities_sf <- uscities_sf |>
  mutate(dist_to_canada_km = set_units(st_distance(geometry, canada_border), "km"))

uscities_sf |>
  st_drop_geometry() |>
  select(city, state_id, dist_to_canada_km) |>
  arrange(desc(dist_to_canada_km)) |>
  slice_head(n = 5) |>
  flextable()

```

# Question 3

## 3.1 Data

```{r}
library(ggplot2)
library(ggrepel)
library(gghighlight)

na <- filter(countries, sovereignt %in% c("Mexico", "Canada", "United States of America")) |>
  st_as_sf()

big_cities <- uscities_sf |> 
  slice_max(population, n = 10)

ggplot()+
  geom_sf(data = na, fill = "grey")+
  geom_sf(data = states, fill = "white", lty = 2, size = 0.5) + 
  geom_sf(data = big_cities, color = "red", size = 0.5) + 
  ggrepel :: geom_label_repel(data = big_cities,
                              aes(geometry = geometry, label = city, label.size = 0.1),
                              stat = "sf_coordinates") +
  theme_void()
```

## 3.2 City Distance to Border

```{r}
# Top 5 farthest cities
# Convert units column to numeric
uscities_sf <- uscities_sf |>
  mutate(dist_to_usa_border_num = as.numeric(dist_to_usa_border_km))

# Get the 5 cities farthest from national border
farthest_from_border <- uscities_sf |>
  slice_max(order_by = dist_to_usa_border_num, n = 5)

# Create label data with x and y coordinates
label_data <- farthest_from_border |>
  mutate(coords = st_coordinates(geometry)) |>
  mutate(x = coords[,1], y = coords[,2]) |>
  st_drop_geometry()

# Plot
ggplot() +
  geom_sf(data = states, fill = NA, color = "gray60") +
  geom_sf(data = uscities_sf, aes(color = dist_to_usa_border_num), size = 0.5) +
  scale_color_viridis_c(option = "plasma", name = "Distance (km)") +
  geom_sf(data = farthest_from_border, color = "red", size = 1.5) +
  ggrepel::geom_label_repel(
    data = label_data,
    aes(x = x, y = y, label = city),
    size = 3
  ) +
  theme_minimal() +
  labs(title = "Distance to National Border (USA)")

```

## 3.3 City Distance from Nearest State

```{r}
uscities_sf <- uscities_sf |>
  mutate(dist_to_state_border_num = as.numeric(dist_to_state_border_km))


# Get top 5 farthest from any state border
farthest_from_state <- uscities_sf |>
  slice_max(order_by = dist_to_state_border_num, n = 5)

# Extract coords for labels
label_state <- farthest_from_state |>
  mutate(coords = st_coordinates(geometry)) |>
  mutate(x = coords[, 1], y = coords[, 2]) |>
  st_drop_geometry()

# Plot
ggplot() +
  geom_sf(data = states, fill = NA, color = "grey70", linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = uscities_sf, aes(color = dist_to_state_border_num), size = 0.6, alpha = 0.7) +
  scale_color_viridis_c(option = "magma", name = "Distance (km)", direction = -1) +
  geom_sf(data = farthest_from_state, color = "#1f77b4", size = 1.5) +
  ggrepel::geom_label_repel(
    data = label_state,
    aes(x = x, y = y, label = city),
    size = 3,
    box.padding = 0.4,
    segment.color = "grey60",
    fill = "white",
    label.size = NA
  ) +
  theme_light(base_size = 13) +
  theme(
    panel.background = element_rect(fill = "#f9f9f9"),
    panel.grid.major = element_line(color = "#e0e0e0", size = 0.3),
    panel.border = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.2, "cm")
  ) +
  labs(
    title = "Distance to Nearest State Border",
    subtitle = "Top 5 most isolated cities from any state line",
    caption = "Source: USAboundaries + uscities.csv"
  )


```

## 3.4 Equidistant Boundary From Mexico and Canada

```{r}
# Create variable for absolute distance difference
uscities_sf <- uscities_sf |>
  mutate(dist_diff = abs(as.numeric(dist_to_mexico_km) - as.numeric(dist_to_canada_km)))

# Filter cities within 100 km of equidistance
equidistant_cities <- uscities_sf |>
  filter(dist_diff <= 100)

# Top 5 most populous in that zone
top5_equidistant <- equidistant_cities |>
  slice_max(order_by = population, n = 5)

ggplot() +
  geom_sf(data = states, fill = NA, color = "gray60") +
  geom_sf(data = uscities_sf, color = "lightgrey", size = 0.3) +
  geom_sf(data = equidistant_cities, color = "green", size = 0.8) +
  gghighlight::gghighlight(dist_diff <= 100, use_direct_label = FALSE) +
  geom_sf(data = top5_equidistant, color = "black", size = 1.5) +
  ggrepel::geom_label_repel(
    data = top5_equidistant |> st_drop_geometry() |> 
      mutate(geometry = st_coordinates(top5_equidistant)),
    aes(x = geometry[, 1], y = geometry[, 2], label = city),
    size = 3
  ) +
  theme_minimal() +
  labs(title = "Cities ~Equidistant (±100 km) from Mexico and Canada")

```

# Question 4

## 4.1 - 100-Mile Border Zone

```{r}
uscities_sf <- uscities_sf |>
  mutate(
    dist_to_usa_border_km = as.numeric(dist_to_usa_border_km),
    in_border_zone = dist_to_usa_border_km <= 160
  )

# Total number of cities
total_cities <- nrow(uscities_sf)

# Cities in 100-mile zone
border_zone_cities <- uscities_sf |>
  filter(in_border_zone)

n_border_cities <- nrow(border_zone_cities)
border_population <- sum(border_zone_cities$population, na.rm = TRUE)
total_population <- sum(uscities_sf$population, na.rm = TRUE)
percent_in_zone <- (border_population / total_population) * 100

library(flextable)

tibble(
  `Total Cities` = total_cities,
  `Cities in Border Zone` = n_border_cities,
  `Population in Border Zone` = border_population,
  `Total Population` = total_population,
  `Percent in Border Zone` = round(percent_in_zone, 1)
) |>
  flextable()

```

## 4.2 Mapping Cities in 100-Mile Zone

```{r}
top10_danger_zone <- border_zone_cities |>
  slice_max(order_by = population, n = 10)

label_data <- top10_danger_zone |>
  mutate(coords = st_coordinates(geometry)) |>
  mutate(x = coords[, 1], y = coords[, 2]) |>
  st_drop_geometry()

ggplot() +
  geom_sf(data = states, fill = NA, color = "grey60") +
  geom_sf(data = uscities_sf, color = "lightgrey", size = 0.3) +
  geom_sf(data = border_zone_cities, aes(color = population), size = 0.6) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  gghighlight::gghighlight(in_border_zone, use_direct_label = FALSE) +
  geom_sf(data = top10_danger_zone, color = "black", size = 1.2) +
  ggrepel::geom_label_repel(
    data = label_data,
    aes(x = x, y = y, label = city),
    size = 3
  ) +
  theme_light() +
  labs(title = "U.S. Cities Within 100 Miles of a National Border",
       subtitle = "Top 10 Most Populous in the ACLU 'Danger Zone'")

```

## Label Most Populated City in Each State

```{r}
# Get top city per state
top_city_per_state <- border_zone_cities |>
  group_by(state_id) |>
  slice_max(order_by = population, n = 1) |>
  ungroup()

label_statewise <- top_city_per_state |>
  mutate(coords = st_coordinates(geometry)) |>
  mutate(x = coords[,1], y = coords[,2]) |>
  st_drop_geometry()

ggplot() +
  geom_sf(data = states, fill = NA, color = "grey60") +
  geom_sf(data = uscities_sf, color = "lightgrey", size = 0.3) +
  geom_sf(data = border_zone_cities, aes(color = population), size = 0.6) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  geom_sf(data = top_city_per_state, color = "black", size = 1.2) +
  ggrepel::geom_label_repel(
    data = label_statewise,
    aes(x = x, y = y, label = city),
    size = 3
  ) +
  theme_light() +
  labs(title = "Most Populous City in Each State within the 100-Mile Danger Zone",
       subtitle = "Based on ACLU's Border Zone Interpretation")

```
